{\bf Topology, Geometry and Morphology of the Dark Matter Web}

### 2018-06-28 Draft/Chapter5/Paper2017b_Nobmath.tex
submitted draft

For a simulation box with $N_p$ particles, each of mass $m_p$, the halo-mass fractions, $\displaystyle f_h = (\sum\limits_{i=1}^{N_H} m_H(i)) / (m_p N_p)$ (where $m_H$ is the mass of each halo and $N_H$ is total number of haloes) are shown in a Venn diagram in \autoref{fig:HaloFinderMF}. For the simulation with $N_p = 128^3$ particles, AHF-, $\lambda_3$-, and FOF- haloes occupy $22$, $31$ and $35$ per cent of the total mass respectively. Nearly $19$ per cent of the total mass are concurrently detected as belonging to haloes by all the three algorithms. FOF (with highest halo mass fraction) detects virtually all the haloes that AHF (with least halo mass fraction). About $3$ per cent of the particles classified as belonging to haloes by both AHF and FOF are not classified as multistream halo particles. Our method also detected nearly $6$ per cent of mass particles as haloes, which neither FOF nor AHF classify as haloes. For simulation with $N_p = 256^3$ particles, the corresponding halo mass fractions $f_h^{\rm AHF} = 30$ per cent, $f_h^{\lambda_3} = 32$ per cent, and $f_h^{\rm FOF} = 42$ per cent respectively. Thus the mass fraction $f_h^{\lambda_3}$ remains fairly consistent over increasing mass resolution, as opposed to AHF and FOF. However, large fractions of these mass particles, nearly $23$ per cent of the $N_p = 256^3$ (increased from $19$ per cent for low mass resolution simulation), are simultaneously detected as belonging to haloes by different methods, as shown in the right panel of \autoref{fig:HaloFinderMF}. For the simulation with $N_p = 256^3$, we also see increase in agreement between any two pairs for halo finders. That is, the mass fraction of haloes simultaneously detected by $\lambda_3$ and FOF jumps from $25$ per cent to $27$ per cent. This correspondence increases from $19$ to $23$ per cent for $\lambda_3$-AHF, and $22$ to $30$ per cent for FOF-AHF pairs. For the same pair (in the $N_p = 256^3$ simulation), $12.3$ per cent of particles are detected by FOF but not AHF, whereas almost all the particles ($>99.9$) for the AHF particles were also detected by FOF. For the pair $\lambda_3$-AHF, $9.5$ per cent of particles are detected by $\lambda_3$ but not AHF, and $7.3$ per cent of particles were detected by AHF but not by $\lambda_3$. Finally, for the $\lambda_3$-FOF pair, $5$ per cent of particles are detected by $\lambda_3$ but not FOF, and $15$ per cent of particles were detected by FOF but not by $\lambda_3$.

>>>>

The halo-mass fractions in a simulation box, $\displaystyle f_h = (\sum\limits_{i=1}^{N_H} m_H(i)) / (m_p N_p)$ (where $N_p$ is the number of particles in the simulation of mass $m_p$ each,  $m_H$ is the mass of each halo and $N_H$ is total number of haloes) are shown in a Venn diagram in \autoref{fig:HaloFinderMF}. For the simulation with $N_p = 128^3$ particles, AHF-, $\lambda_3$-, and FOF- haloes occupy $22$, $31$ and $35$ per cent of the total mass respectively. Nearly $19$ per cent of the total mass are concurrently detected as belonging to haloes by all the three algorithms. FOF (with highest halo mass fraction) detects virtually all the haloes that AHF (with least halo mass fraction). About $3$ per cent of the particles classified as belonging to haloes by both AHF and FOF are not classified as multistream halo particles. Our method also detected nearly $6$ per cent of mass particles as haloes, which neither FOF nor AHF classify as haloes. For simulation with $N_p = 256^3$ particles, the corresponding halo mass fractions $f_h^{\rm AHF} = 30$ per cent, $f_h^{\lambda_3} = 32$ per cent, and $f_h^{\rm FOF} = 42$ per cent respectively. Thus the mass fraction $f_h^{\lambda_3}$ remains fairly consistent over increasing mass resolution, as opposed to AHF and FOF. However, large fractions of these mass particles, nearly $23$ per cent of the $N_p = 256^3$ (increased from $19$ per cent for low mass resolution simulation), are simultaneously detected as belonging to haloes by different methods, as shown in the right panel of \autoref{fig:HaloFinderMF}. For the simulation with $N_p = 256^3$, we also see increase in agreement between any two pairs for halo finders. That is, the mass fraction of haloes simultaneously detected by $\lambda_3$ and FOF jumps from $25$ per cent to $27$ per cent. This correspondence increases from $19$ to $23$ per cent for $\lambda_3$-AHF, and $22$ to $30$ per cent for FOF-AHF pairs. For the same pair (in the $N_p = 256^3$ simulation), $12.3$ per cent of particles are detected by FOF but not AHF, whereas almost all the particles ($>99.9$) for the AHF particles were also detected by FOF. For the pair $\lambda_3$-AHF, $9.5$ per cent of particles are detected by $\lambda_3$ but not AHF, and $7.3$ per cent of particles were detected by AHF but not by $\lambda_3$. Finally, for the $\lambda_3$-FOF pair, $5$ per cent of particles are detected by $\lambda_3$ but not FOF, and $15$ per cent of particles were detected by FOF but not by $\lambda_3$.

### 2018-06-28 Draft/thesis-ku.tex
submitted draft

% Set the title and author info
\title{Topology, geometry and morphology of the dark matter web}
\author{Nesar Ramachandra}

\dept{Department of Physics and Astronomy}
\degreetitle{Doctor of Philosophy}
\papertype{Dissertation} %capitalization is important here
\committee{Sergei Shandarin}{Barbary Anthony-Twarog}{Hume Feldman}{Suzanne Shontz}{Bruce Twarog}
%AT Most 5 members allowed, last here is blank on purpose to demonstrate
%flexibility

@@

Spatial distribution of dark matter displays variety of intricate three dimensional structures on the largest scales in the Universe, notably the massive haloes, long tubular filaments, flattened walls and the vast under-dense voids. Galaxies embedded in the dark matter structures have illuminated rich geometry of these structures currently known as the cosmic web.

Cosmological N-body simulations are indispensable tools for understanding the evolution of the dark matter web. Recent developments in the numerical analysis of cosmological simulations have hinted towards incorporating the dynamical information of gravitational clustering of collisionless dark matter. This is inferred from a six-dimensional Lagrangian sub-manifold -- comprising of initial and final coordinates of dark matter particles. Velocity multistream field derived from this sub-manifold sheds new light on the nature of gravitational collapse.

Geometrical, topological, morphological and heuristic diagnostic tools used in this novel parameter space reveal features of the dark matter distribution. For instance, a single void structure not only percolates the multistream field in all the directions, but also occupies over 99 per cent of all the single-streaming regions. On the other hand, connected filaments display a rapid topological transition to isolated islands of high multistream values. Hessian analysis delineates structures with different shapes: tubular, sheet-like, or globular -- enabling detection of dark matter haloes without ad hoc parameters related to matter density or distance field.

@@

I would also like to thank my thesis committee: Prof. Hume Feldman, Prof. Bruce Twarog, Prof. Barbara Anthony-Twarog, Dr. Suzanne Shontz, Prof. Mikhail Medvedev for supervision and guidance that has been important for the materialization of this thesis.

My appreciation to the friendly faculty and staff at the Department of Physics and Astronomy, especially Kristin Rennells, Kim Hubbel and Joel Sauerwein whose readiness to help students cannot be overestimated.

My time as a graduate student was enjoyable in large part due to many wonderful people I was surrounded by. Anupama has been an inspiring grad student and a wonderful friend, with whom I have shared moments of frustration, laughter and excitement through the entirety of my PhD. I had a great pleasure sharing offices with Brett, Alex, Yuyu and Keita over the past few years. Stimulating conversations with them -- physics or otherwise -- was always fun and exciting. Special thanks to Donald and Sinan for the company that kept me sane over the course of PhD. I've had amazing house mates: Niranjan, Niharika, Darshan and Charu, who have not only put up with my idiosyncrasies, but also occasionally cooked me delicious dinner.

I very much appreciate the opportunities provided by Prof. Salman Habib and the Cosmology group at Argonne, where I have spent better part of this past year. I'm really excited about joining the group as a post-doc in the near future.

Thank you Mum and Dad, for unconditional love and support throughout graduate school and my life in general. They have always encouraged me to pursue my dreams. None of this would have been possible without their care and guidance.

>>>>

% Set the title and author info
\title{ {\bf Topology, Geometry and Morphology of the Dark Matter Web} }
\author{Nesar Ramachandra}

\dept{Department of Physics and Astronomy}
\degreetitle{{ \it Doctor of Philosophy} }
\papertype{Dissertation} %capitalization is important here
\committee{Sergei Shandarin}{Barbary Anthony-Twarog}{Hume Feldman}{Suzanne Shontz}{Bruce Twarog}
%AT Most 5 members allowed, last here is blank on purpose to demonstrate
%flexibility

@@

Spatial distribution of dark matter displays a variety of intricate three dimensional structures on the largest scales in the Universe, notably the massive haloes, long tubular filaments, flattened walls and the vast under-dense voids. Galaxies embedded in the dark matter structures have illuminated the rich geometry of these structures currently known as the cosmic web.

Cosmological N-body simulations are indispensable tools for understanding the evolution of the dark matter web. Recent developments in the numerical analysis of these simulations have hinted towards incorporating the dynamical information of gravitational clustering of collisionless dark matter. This is inferred from a six-dimensional Lagrangian sub-manifold -- comprising of initial and final coordinates of the dark matter particles. Velocity multistream field derived from this sub-manifold sheds new light on the nature of gravitational collapse.

Geometrical, topological, morphological and heuristic diagnostic tools used in this novel parameter space reveal features of the dark matter distribution. For instance, a single void structure not only percolates the multistream field in all the directions, but also occupies over 99 per cent of all the single-streaming regions. On the other hand, connected filaments display a rapid topological transition to isolated islands at high multistream values. Hessian analysis delineates structures with different shapes: tubular, sheet-like, or globular -- enabling detection of the dark matter haloes without ad hoc parameters related to matter density or distance field.

@@

\bigskip
I would also like to thank my thesis committee: Prof. Hume Feldman, Prof. Bruce Twarog, Prof. Barbara Anthony-Twarog, Dr. Suzanne Shontz, Prof. Mikhail Medvedev for supervision and guidance that has been important for the materialization of this thesis.

\bigskip
My appreciation to the friendly faculty and staff at the Department of Physics and Astronomy, especially Kristin Rennells, Kim Hubbel and Joel Sauerwein whose readiness to help students cannot be overestimated.

\bigskip
My time as a graduate student was enjoyable in large part due to many wonderful people I was surrounded by. Anupama has been an inspiring grad student and a wonderful friend, with whom I have shared moments of frustration, laughter and excitement through the entirety of my PhD. I had a great pleasure sharing offices with Brett, Alex, Yuyu and Keita over the past few years. Stimulating conversations with them -- physics or otherwise -- was always fun and exciting. Special thanks to Donald and Sinan for the company that kept me sane over the course of PhD. I've had amazing house mates: Niranjan, Niharika, Darshan and Charu, who have not only put up with my idiosyncrasies, but also occasionally cooked me delicious dinner.

\bigskip
I very much appreciate the opportunities provided by Prof. Salman Habib and the Cosmology group at Argonne, where I have spent better part of this past year. I'm really excited about joining the group as a post-doc in the near future.

\bigskip
Thank you Mum and Dad, for unconditional love and support throughout graduate school and my life in general. They have always encouraged me to pursue my dreams. None of this would have been possible without their care and guidance.
